---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
  
Load needed libraries
*dplyr
*lattice
```{r, echo=TRUE}
library(dplyr)
library(lattice)
```
  
## Loading and preprocessing the data
Unzip activity source .csv file and read it into the activity data frame.  
Ensure the second column is treated as a Date type going forward.
```{r, echo=TRUE}
activity_data <- read.csv("activity.csv", header=TRUE, colClasses=c(NA,"Date",NA))
str(activity_data)
```
  
## What is mean total number of steps taken per day?
Show the step frequency counts of total steps by date.
```{r, echo=TRUE}
summary_by_date <- activity_data %>% 
    group_by(date) %>%
    summarize(totalSteps = sum(steps,na.rm=TRUE)) 

hist(summary_by_date$totalSteps, main = "Histogram of Total Steps", xlab="Total Steps", breaks=10)

```
  
The mean and median of the total steps are:
```{r, echo=TRUE}
mean(summary_by_date$totalSteps)
median(summary_by_date$totalSteps)
```
  
## What is the average daily activity pattern?
```{r, echo=TRUE}
average_by_5min_interval <- activity_data %>% 
    group_by(interval) %>%
    summarize(meanSteps = mean(steps,na.rm=TRUE)) 

plot(average_by_5min_interval, type="l", xlab= "Interval", ylab= "Average # of Steps", col="green" , lwd=2)
```
  
Determine which 5 minute interval has the maximum steps on average.
```{r, echo=TRUE}
average_by_5min_interval %>% filter(meanSteps == max(meanSteps))
```
  
## Imputing missing values  
  
The total number of missing values in the source dataset is:
```{r, echo=TRUE}
nrow(activity_data[is.na(activity_data$steps),])
```
  
Impute missing values from the mean # of steps for that 5 minute interval.
A re-sort of the data is needed after the merge.
```{r, echo=TRUE}
filled_activity_data <- merge(activity_data, average_by_5min_interval, by="interval")
filled_activity_data <- filled_activity_data %>% 
    mutate(steps = ifelse(is.na(steps), meanSteps, steps) ) %>% 
    select(steps,date,interval) %>%
    arrange(date,interval)
```
  
Show the step frequency counts by date with imputed values
```{r, echo=TRUE}
filled_sum_steps_by_date <- filled_activity_data %>% 
    group_by(date) %>%
    summarize(totalSteps = sum(steps,na.rm=TRUE)) 

hist(filled_sum_steps_by_date$totalSteps, main = "Histogram of Total Steps (Imputed)", xlab="Total Steps", breaks=10)
```
  
```{r, echo=TRUE}
mean(filled_sum_steps_by_date$totalSteps)
median(filled_sum_steps_by_date$totalSteps)
```
The impact of using imputed values has increased the average number of steps per day as well as the median. 
  
## Are there differences in activity patterns between weekdays and weekends?
First we add a new factor for whether the date is on a weekday or a weekend.
Then we'll compare the two date types using a panel time series plot of average # of steps taken by 5 minute interval.
```{r, echo=TRUE}
weekday_activity_data <- mutate(filled_activity_data, 
     weekdayType = ifelse(weekdays(filled_activity_data$date) %in% c("Saturday","Sunday"),
                           "weekend", "weekday")
                          )

average_by_5min_interval_with_weekday <- weekday_activity_data %>% 
    group_by(interval, weekdayType) %>%
    summarize(averageSteps = mean(steps,na.rm=TRUE)) 

xyplot(averageSteps ~ interval | weekdayType,
                data = average_by_5min_interval_with_weekday,
                type = "l",
                lty = c(1, 2, 2, 1),
                lwd = c(1, 1, 1, 3),
                col.line = c(rep("black",3), "red"))

```
  
From the chart we can see that on weekdays most steps are taken in the morning around 8:00am,  
while on weekends the steps are spread out across the daytime.

